---
title: "forest_cover_type"
author: "Ada Kiekhaefer"
date: "3/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Classification of forest cover type
##Introduction
The actual forest cover type for a given observation (30 x 30 meter cell) was determined from US Forest Service (USFS) Region 2 Resource Information System (RIS) data. The study area included four wilderness areas found in the Roosevelt National Forest of northern Colorado. 
Some background information for these four wilderness areas:  
	Neota (area 2) probably has the highest mean elevational value of 
	the 4 wilderness areas. Rawah (area 1) and Comanche Peak (area 3) 
	would have a lower mean elevational value, while Cache la Poudre 
	(area 4) would have the lowest mean elevational value. 
As for primary major tree species in these areas, Neota would have 
	spruce/fir (type 1), while Rawah and Comanche Peak would probably
	have lodgepole pine (type 2) as their primary species, followed by 
	spruce/fir and aspen (type 5). Cache la Poudre would tend to have 
	Ponderosa pine (type 3), Douglas-fir (type 6), and 
	cottonwood/willow (type 4).

### List of variables

## Exploratory Data Analysis
###Loading r libraries 
``` {r load_library}
#Loading r libraries
library(dplyr)
library(readr)
```

### Loading data
* The data file was in csv format. I used read_csv function from readr package to load it into R. 
* Unfortuanely, none of the columns were labeled. 

```{r read_data}
#importing data
df_raw <- read_csv('./data/covtype.data', col_names=FALSE)
#Looking at the first few rows of the data
df_raw %>% head()
```

### Added column names
Added column names based on the information in covtype.info. 

``` {r add_column_names}
temp_name <- c('Elevation', 'Aspect', 'Slope', 'Horizontal_Distance_To_Hydrology', 'Vertical_Distance_To_Hydrology', 
             'Horizontal_Distance_To_Roadways', 'Hillshade_9am', 'Hillshade_Noon', 'Hillshade_3pm',
             'Horizontal_Distance_To_Fire_Points', 'Wilderness_Area1', 'Wilderness_Area2', 'Wilderness_Area3',
             'Wilderness_Area4')

# create soil column names
num <-1:40
soil_name <- paste('Soil_Type', num, sep='')
col_names <- c(temp_name, soil_name, 'Cover_Type')

# add column names to the dataframe
df <- df_raw
colnames(df) <- col_names
```
### Checked data structures
* There are 581,012 observations and 55 variables in the dataset including the response variable. 
* Values in columns are all numeric.  
* Some features should be categorical (wilderness area, soil type)
* Wilderness_Area variable was recorded as 4 binary variables
* Soil_Type variable was recorded as 40 binary variables 

``` {r check_data}
df_raw %>% glimpse()
```

### Check if there is any missing value
* No missing value in this dataset. 
```{r check_NA}
any(is.na(df))
```



## 1. Describe the dataset and any interesting variables (feel free to include response).
Hint: It may be useful to provide a high-level summary (i.e. descriptive statistics) on the variables

* Features with different scales:  
* Elevation: all cover types are at quite high elevation with minimum elevation at 1859 meters. 
* Slope: range of slope is quite large with minimum 0 and maximum 66 degree (high slope, mountainous areas)
* Wilderness_Area: mostly in wilderness area 1 (mean 0.4489) and 3 (mean 0.4361)
* Cover_Type: the response variable has type 1 with the ratio of 0.365 and and type 2 with the ratio of 0.488 (see code below)

``` {r summary_statistics}
#summary statistics of all features
summary(df)

# standard deviations of numerical features
sd_numeric <- df %>% 
  select(Elevation:Horizontal_Distance_To_Fire_Points) %>%
  summarise_each(funs(sd))
print(sd_numeric)

#ratio of Cover_Type
df %>% 
  group_by(Cover_Type) %>%
  summarise(n=n()) %>%
  mutate(freq = n/sum(n))
```

```{r correlation}
#correlation of continuous features
#subset dataframe, remove categorical variables
df_sub <- df %>%
  select(-starts_with("Soil"), -starts_with("Wild"), -'Cover_Type')

#calculate correlations among numeric features
df_sub_corr <- df_sub %>% 
  cor() 

#loading library for correlation plot
library(ggcorrplot)

#correlation plot
df_sub_corr %>% ggcorrplot(type = "lower", outline.col = "white")
```

## boxplot 



## Feature Engineering


## Model Straegy
##Multinomial Logistic Regression


## Results

